The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-23 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revisions, mainly on section 5 and the appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2586 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b7890df49716d07a5e3ffb8c27c221676de1855a

\begin{table}
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$\\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{\label{tab:variables}Variables included in the data generating process 
as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_i(Z_1,Z_2)$, $i=1,2$.}
\end{table}

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
can depend on the true split variable $Z_1$. More specifically, three different variations
are considered for the coefficients $\beta_0$ and $\beta_1$:
%
\begin{enumerate}
  \item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
  \item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
        (at $0$).
  \item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}
%
See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
changes ($\beta_j$ with $j\in\{0,1\}$) switches between two values at the split point $\xi$:
%
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \leq \xi \\
+\delta \cdot (-1)^{j}  \quad \text{if } Z_1 > \xi
\end{cases}
\end{align*}
%
%while in the continuous case a simple linear function is considered in the form of
%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.

@@

\subsubsection{``Tree'' scenario}
For the second simulation scenario (``tree'' scenario) the same basic setting with the same 
variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
However, not only $Z_1$ but also $Z_2$ is used as a true split variable. Therefore, an additional split
is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
to a change in the intercept $\beta_0$.
Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
%
\begin{align*}
\beta_{0}(Z_1, Z_2) = \begin{cases}
0       \qquad \text{if } Z_2 \leq \xi \\
-\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
+\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
\end{cases}
\end{align*}
\begin{align*}
\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
+\delta \quad \text{if } Z_2 \leq \xi \\
-\delta \quad \text{if } Z_2 > \xi
\end{cases}
\end{align*}
%
The underlying tree structure for the data generating process is illustrated in 
Figure~\ref{fig:dgp_tree}.

@@

Moreover, as defined by the true tree structure used in the data generating process, the
parameter functions $\beta_i(Z_1)$, $i=1,2$, are step functions which is particularly 
advantageous for the maximum selection in the test statistic applied by MOB while the linear 
selection in CTree is preferable for detecting monotonous dependencies. To show that employing 
a continuous parameter function instead of a step function only has an impact on whether CTree or
MOB performs better the exact same simulation setting as used for Figure~\ref{fig:scores_stump} 
but with continuously changing parameters has been evaluated and is illustrated in 
Appendix~\ref{app:contbeta}.

@@

For a central split point ($\xi=0$ corresponding to the median, upper panels) all strategies 
perform similarly well as long as the intercept varies (left and right panel) with CTree and 
MOB being only slightly ahead. However, for the split point at the $90$\% quantile $0.8$ 
(lower panels) the performance of all of the applied strategies decreases. 
While both GUIDE versions struggle to detect the correct split variable even for a high effect 
size, MOB is clearly ahead with CTree leading to the second best results.
As discussed before, this advantage of MOB over CTree is mainly due to the abrupt shift in the 
model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
continuously changing parameters (see Appendix~\ref{app:contbeta}).
From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
can be deduced that categorizing split variables and binarizing scores has a negative effect 
on the ability to detect the right split variable. To elaborate these effects more in detail 
the results for all possible combinations of the building blocks are discussed in a full 
factorial 3-way analysis in the following section.

\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
To investigate the effects of each of the building blocks separately the most general 
case of the ``stump'' scenario where the intercept and 
slope parameter are both varying is considered. In the underlying evaluation all 
possible combinations of the building blocks have been included. The different levels of 
each of the three building blocks are listed in Table~\ref{tab:levels} where 
$h$ and $g$ refer to the transformation functions applied to the response $Y$ or
a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
(See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
for a more detailed overview of all 12 combinations.)
If quartiles are used as breakpoints to categorize a split variable the employed
function $g$ maps to a four-dimensional vector of zeros except for one value $1$
indication the bin to which an observation is assigned to. For a maximum selection
of a split variable $g$ maps to a vector containing only zeros until a certain
changepoint from which on the values are replace by $1$. This breakpoint indicates
for example a crossed quantile. In linear selection $g$ is simply the identity function.

@@

Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing 
residuals/scores decreases the performance as it leads to higher $p$-values corresponding 
to the true split variable $Z_1$ and thus to lower power in the presented settings. 
For a central true split point ($\xi=0$, left panel) this effect remains at an almost constant 
level regardless of whether possible split variables are categorized or not. However, in case 
of the true split point at the $90$\% quantile $0.8$ (right panel), categorizing split 
variables increases the $p$-value of $Z_1$ even more. 
Here, the maximum selection performs best while the linear selection shows its disadvantage in 
detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximum
selection of a split variable is advantageous. But unless one of the breakpoints used for binning 
possible split variables is close to the true split point (as for the median $0$, left panel),
this categorization weakens the performance of a testing strategy. 
While the effect of both of these transformations of the data (binarization of 
residuals/scores and categorization of split variables) can be observed
separately in Figure~\ref{fig:3way}, combining them increases the negative impact 
remarkably.

@@

For MOB and CTree it has been shown that the tests work very well as significance tests.
Therefore, it can be recommended to take advantage of this power and use the information
given by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:tree_all}
in most cases the results of these two testing strategies do not vary significantly over 
a change of the applied pruning strategy.
On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
strategy should not be based on their results which argues for the application of post-pruning.
This is also supported by the results in Figure~\ref{fig:tree_all} showing that post-pruning improves
the performance of both GUIDE versions. Particularly if the true split point is close to a binning 
point (such that categorizing split variables has no negative effect) the post-pruning strategy 
compensates the lack of power of GUIDE+scores compared to CTree and MOB almost completely (bottom 
row, first and third panel).
In conclusion, the choice of a pruning strategy should be based on knowledge about the 
power of the applied testing strategy. With a well working significance test additional expensive
calculations such as needed for post-pruning may be avoided being one of the main advantages of 
pre-pruning. However, deficiencies of a testing strategy in terms of power can be at least partly 
evened out by employing a post-pruning strategy.

@@

Also in more general models residual-based and score-based procedures are expected to perform equally 
well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
that are not connected to the mean might be missed when only considering residuals. This can be prevented 
by including scores which are a standard measurement of goodness of fit that can be easily employed for 
all model types.
Even though it also depends on the type of the applied residuals 
(see, e.g., \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002} 
and \citealp{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of of testing all parameters.

@@

The functions \code{ctree} and \code{mob} provide an implementation of the
two tree algorithms in their original form. For their adapted versions additional
modifications have been applied within these functions allowing for a categorization
of possible split variables and a binarization of scores.
To evaluate the GUIDE algorithm a reimplementation of this algorithm has been 
established/constructed using the basic framework of \code{ctree} and \code{mob}.

@@

\section{Test statistics}
This section provides further information on the test statistics applied in CTree, MOB, 
and GUIDE. The presented details have been extracted from the original papers 
\cite{Hothorn+Hornik+Zeileis:2006}, \cite{Zeileis+Hothorn+Hornik:2008}, 
and \cite{Loh:2002}, however, notation is adapted to the main manuscript
in order to allow for better comparison.

\subsection{CTree}
To measure the association of a response $Y$ and each possible split variable $Z$ the
CTree algorithm applies a linear test statistic $T$ which is excerpted from Section~3.1.
(``Variable selection and stopping criteria'') of the original paper \citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
$$
T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i, (Y_1,\ldots,Y_n))^{\top}\right)
$$
where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom transformation of the split variable $Z$ and the influence function $h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the responses $(Y_1, \ldots, Y_n)$ in a permutation symmetric way. Moreover, by applying the ``vec'' operator the resulting $p~\times~q$ matrix is converted into a $p q$ column vector.
Following \cite{Strasser+Weber:1999}
the conditional expectation $\mu$ and covariance $\Sigma$ of a linear test statistic can 
be calculated and used to standardize an observed linear test statistic $t$ within a 
function $c$ mapping into the real line.
For example, the maximum of the absolute values of the standardized linear statistic
$$
c_{\max}(t,\mu,\Sigma) = \max_{k=1,\ldots,pq} \left| \frac{(t-\mu)}{\sqrt{(\Sigma)_{kk}}} \right|
$$
or a quadratic form
$$
c_{\text{quad}}(t, \mu, \Sigma) = (t - \mu)\Sigma^{+} (t - \mu)^{\top}
$$
can be considered where $\Sigma^{+}$ is the Moore-Penrose inverse of $\Sigma$.
Since the asymptotic conditional distribution of a linear test statistic $T$ is a 
multivariate normal with parameters $\mu$ and $\Sigma$ \citep{Strasser+Weber:1999}, 
the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
the corresponding $p$-values can be calculated easily.

@@

For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks are listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``+res'' refers to using only residuals (instead of all scores),
``+scores'' to using all scores (instead of only residuals),
``+cat'' to applying a categorization of split variables,
``+bin'' to applying a binarization of scores/residuals.

\begin{table}[h!]
\centering
\begin{tabular}{ l l l l }
\hline 
Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
\hline
GUIDE             & residuals          & bin          & cat \\
MOB+bin+res     & residuals          & bin          & max \\
CTree+bin+res   & residuals          & bin          & lin \\
MOB+cat+res    & residuals          & lin          & cat \\
MOB+res          & residuals          & lin          & max \\
CTree+res        & residuals          & lin          & lin \\
GUIDE+scores     & scores             & bin          & cat \\
MOB+bin          & scores             & bin          & max \\
CTree+bin        & scores             & bin          & lin \\
MOB+cat          & scores             & lin          & cat \\
MOB               & scores             & lin          & max \\
CTree             & scores             & lin          & lin \\
\hline
\end{tabular}
\caption{Testing strategies applied (representing the 12 combinations
of building blocks).
\label{tab:combinations_appendix}}
\end{table}

@@

\begin{figure}[t]
\setkeys{Gin}{width=\linewidth}
<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "ctree_max", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree", "CTree+max", "MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], 
                                             fill = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")]))

library("latticeExtra")
useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "CTree+max", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
                      par.settings = par.settings),

@@

@
\caption{\label{fig:scores_stump_wosign}
Proportion of replications where the true split variable $Z_1$ shows the smallest $p$-value, 
hence, where $Z_1$ is selected without any significance level in the ``stump'' scenario.
The testing strategies CTree, CTree+max, MOB, GUIDE, and GUIDE+scores are evaluated 
over increasing effect size $\delta$ with 100 replications of 250 observations per step. 
The true split point in $Z_1$ is either the median 0 (top) or the 90\% quantile 0.8 (bottom)
for either varying intercept $\beta_0$ (left), varying slope $\beta_1$
(middle), or both coefficients varying (right).}
\end{figure}

In Section~\ref{sec:scores} the investigated testing strategies are compared based
on the number of replications in which the true split variable $Z_1$ is detected with a $p$-value
smaller than the predefined significance level $\alpha = 0.05$. This choice of measurement
is due to the aim of investigating how well the tests perform as significance test.
However, ignoring this significance level yields the same conclusions 
which can be seen when comparing Figures~\ref{fig:scores_stump_wosign}
and \ref{fig:scores_stump}.
In Figure~\ref{fig:scores_stump_wosign} the proportion of replications 
in which the true splitting variable $Z_1$ is detected, but not necessarily with a $p$-value 
smaller than the significance level, is illustrated. In Figure~\ref{fig:scores_stump} 
the exact same simulation setting is applied with a significance level.

>>>>

\begin{table}
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$\\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{\label{tab:variables}Variables included in the data generating process 
as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_j(Z_1,Z_2)$, $j=0,1$.}
\end{table}

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
can depend on the true split variable $Z_1$. More specifically, three different variations
are considered for the coefficients $\beta_0$ and $\beta_1$:
%
\begin{enumerate}
  \item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
  \item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
        (at $0$).
  \item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}
%
See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
changes ($\beta_j$ with $j\in\{0,1\}$) switches between two values at the split point $\xi$:
%
\begin{align*}
\beta_{j}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \leq \xi \\
+\delta \cdot (-1)^{j}  \quad \text{if } Z_1 > \xi
\end{cases}
\end{align*}
%
%while in the continuous case a simple linear function is considered in the form of
%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.

@@

\subsubsection{``Tree'' scenario}
For the second simulation scenario (``tree'' scenario) the same basic setting with the same 
variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
However, not only $Z_1$ but also $Z_2$ is used as a true split variable following a uniform distribution
on $[-1,1]$. Therefore, an additional split
is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
to a change in the intercept $\beta_0$.
Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
%
\begin{align*}
\beta_{0}(Z_1, Z_2) = \begin{cases}
0       \qquad \text{if } Z_2 \leq \xi \\
-\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
+\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
\end{cases}
\end{align*}
\begin{align*}
\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
+\delta \quad \text{if } Z_2 \leq \xi \\
-\delta \quad \text{if } Z_2 > \xi
\end{cases}
\end{align*}
%
The underlying tree structure for the data generating process is illustrated in 
Figure~\ref{fig:dgp_tree}.

@@

Moreover, as defined by the true tree structure used in the data generating process, the
parameter functions $\beta_j(Z_1)$, $j=0,1$, are step functions which is particularly 
advantageous for the maximum selection in the test statistic applied by MOB while the linear 
selection in CTree is preferable for detecting monotonous dependencies. To show that employing 
a continuous parameter function instead of a step function only has an impact on whether CTree or
MOB performs better the exact same simulation setting as used for Figure~\ref{fig:scores_stump} 
but with continuously changing parameters has been evaluated and is illustrated in 
Appendix~\ref{app:contbeta}.

@@

For a central split point ($\xi=0$ corresponding to the median, upper panels) all strategies 
perform similarly well as long as the intercept varies (left and right panel) with CTree and 
MOB being only slightly ahead. However, for the split point at the $90$\% quantile $0.8$ 
(lower panels) the performance of all of the applied strategies decreases. 
While both GUIDE versions struggle to detect the correct split variable even for a high effect 
size, MOB is clearly ahead with CTree leading to the second best results.
As discussed before, this advantage of MOB over CTree is mainly due to the abrupt shift in the 
model parameters $\beta_0$ and $\beta_1$ and turns into an advantage of CTree over MOB for 
continuously changing parameters (see Appendix~\ref{app:contbeta}).
From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
can be deduced that categorizing split variables and binarizing scores also decreases the
ability to detect the right split variable. To elaborate these effects more in detail 
the results for all possible combinations of the building blocks are discussed in a full 
factorial 3-way analysis in the following section.

\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
To investigate the impact of each of the building blocks separately the most general 
case of the ``stump'' scenario where the intercept and slope parameter are both varying is 
considered. In the underlying evaluation all possible combinations of the building blocks 
have been included. The different levels of each of the three building blocks are listed in 
Table~\ref{tab:levels} where $h$ and $g$ refer to the transformation functions applied to the 
response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
If quartiles are used as breakpoints to categorize a split variable the employed
function $g$ maps to a four-dimensional vector of zeros except for one value $1$
indication the bin to which an observation is assigned. For a maximum selection
of a split variable $g$ maps to a vector containing only zeros until a certain
changepoint from which on the values are replaced by $1$. This breakpoint indicates
for example a crossed quantile. In a linear selection strategy $g$ is simply the identity function.
(See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
for a more detailed overview of all 12 combinations.)

@@

Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing 
residuals/scores decreases the performance as it leads to higher $p$-values corresponding 
to the true split variable $Z_1$ and thus to lower power in the presented settings. 
For a central true split point ($\xi=0$, left panel) this effect remains at an almost constant 
level regardless of whether possible split variables are categorized or not. However, in case 
of the true split point at the $90$\% quantile $0.8$ (right panel), categorizing split 
variables increases the $p$-value of $Z_1$ even more. 
Here, the maximum selection performs best while the linear selection shows its disadvantage in 
detecting abrupt changes, but overall it depends on the situation whether a linear or a maximum
selection of a split variable is advantageous. 
Comparing the two panels suggests that unless one of the breakpoints used for binning 
possible split variables is close to the true split point (as for the median $0$, left panel), 
a categorization weakens the performance of a testing strategy. 
Moreover, while the effect of both of these transformations of the data (binarization of 
residuals/scores and categorization of split variables) can be observed
separately in Figure~\ref{fig:3way}, combining them increases the negative impact 
remarkably.

@@

For MOB and CTree it has been shown that their tests work very well as significance tests.
Therefore, it can be recommended to take advantage of this power and use the information
provided by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:tree_all}
in most cases the results of these two testing strategies do not vary significantly over 
a change of the applied pruning strategy.
On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
strategy should not be based on their results which argues for the application of post-pruning.
This is also supported by the results in Figure~\ref{fig:tree_all} showing that post-pruning improves
the performance of both GUIDE versions. Particularly if the true split point is close to a binning 
point (such that categorizing split variables has no negative effect) the post-pruning strategy 
compensates the lack of power of GUIDE+scores compared to CTree and MOB almost completely (bottom 
row, first and third panel).
In conclusion, the choice of a pruning strategy should be based on knowledge about the 
power of the applied testing strategy. With a well working significance test additional expensive
calculations such as needed for post-pruning may be avoided being one of the main advantages of 
pre-pruning. However, deficiencies of a testing strategy in terms of power can be at least partly 
evened out by employing a post-pruning strategy.

@@

Also in more general models residual-based and score-based procedures are expected to perform equally 
well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
that are not connected to the mean might be missed when only considering residuals. This can be prevented 
by including scores which are a standard measurement of goodness of fit that can be easily employed for 
all model types.
Even though it also depends on the type of the applied residuals 
(see, e.g., \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002} 
and \citealp{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of testing all parameters.

@@

The functions \code{ctree} and \code{mob} provide an implementation of the
two tree algorithms in their original form. For their adapted versions additional
modifications have been applied within these functions allowing for a categorization
of possible split variables and a binarization of scores.
To evaluate the GUIDE algorithm a reimplementation of this algorithm has been 
built using the basic framework of \code{ctree} and \code{mob}.

@@

\section{Test statistics}
This section provides further information on the test statistics applied in CTree, MOB, 
and GUIDE. Most of the presented details have been extracted from the original papers 
\cite{Hothorn+Hornik+Zeileis:2006}, \cite{Zeileis+Hothorn+Hornik:2008}, 
and \cite{Loh:2002}, however, notation is adapted to the main manuscript
in order to allow for better comparison.

\subsection{CTree}
To measure the association of a response $Y$ and each possible split variable $Z$ the
CTree algorithm applies a linear test statistic $T$ which is excerpted from Section~3.1.
(``Variable selection and stopping criteria'') of the original paper \citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
$$
T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i, (Y_1,\ldots,Y_n))^{\top}\right)
$$
where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom transformation of the split variable $Z$ and the influence function $h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the response $Y = (Y_1, \ldots, Y_n)$ in a permutation symmetric way. Moreover, by applying the ``vec'' operator the resulting $p~\times~q$ matrix is converted into a $p q$ column vector.
Following \cite{Strasser+Weber:1999}
the conditional expectation $\mu$ and covariance $\Sigma$ of a linear test statistic can 
be calculated and used to standardize an observed linear test statistic $t$ within a 
function $c$ mapping into the real line.
For example, the maximum of the absolute values of the standardized linear statistic
$$
c_{\max}(t,\mu,\Sigma) = \max_{k=1,\ldots,pq} \left| \frac{(t-\mu)}{\sqrt{(\Sigma)_{kk}}} \right|
$$
or a quadratic form
$$
c_{\text{quad}}(t, \mu, \Sigma) = (t - \mu)\Sigma^{+} (t - \mu)^{\top}
$$
can be considered where $\Sigma^{+}$ is the Moore-Penrose inverse of $\Sigma$.
Since the asymptotic conditional distribution of a linear test statistic $T$ is a 
multivariate normal with parameters $\mu$ and $\Sigma$ \citep{Strasser+Weber:1999}, 
the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
the corresponding $p$-values can be calculated easily.

@@

For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks are listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``+res'' refers to using only residuals (instead of all model scores),
``+scores'' to using all model scores (instead of only residuals),
``+cat'' to applying a categorization of split variables,
``+bin'' to applying a binarization of scores/residuals.

\begin{table}[h!]
\centering
\begin{tabular}{ l l l l }
\hline 
Testing Strategy  & Residuals/Scores   & Binarization & Categorization \\
\hline
GUIDE             & residuals          & yes          & cat \\
MOB+bin+res       & residuals          & yes          & max \\
CTree+bin+res     & residuals          & yes          & lin \\
MOB+cat+res       & residuals          & no           & cat \\
MOB+res           & residuals          & no           & max \\
CTree+res         & residuals          & no           & lin \\
GUIDE+scores      & scores             & yes          & cat \\
MOB+bin           & scores             & yes          & max \\
CTree+bin         & scores             & yes          & lin \\
MOB+cat           & scores             & no           & cat \\
MOB               & scores             & no           & max \\
CTree             & scores             & no           & lin \\
\hline
\end{tabular}
\caption{Testing strategies applied (representing the 12 combinations
of building blocks).
\label{tab:combinations_appendix}}
\end{table}

@@

\begin{figure}[t]
\setkeys{Gin}{width=\linewidth}
<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree", "MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

library("latticeExtra")
useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
                      par.settings = par.settings),

@@

@
\caption{\label{fig:scores_stump_wosign}
Proportion of replications where the true split variable $Z_1$ shows the smallest $p$-value, 
hence, where $Z_1$ is selected without any significance level in the ``stump'' scenario.
The testing strategies CTree, MOB, GUIDE, and GUIDE+scores are evaluated 
over increasing effect size $\delta$ with 100 replications of 250 observations per step. 
The true split point in $Z_1$ is either the median 0 (top) or the 90\% quantile 0.8 (bottom)
for either varying intercept $\beta_0$ (left), varying slope $\beta_1$
(middle), or both coefficients varying (right).}
\end{figure}

In Section~\ref{sec:scores} the investigated testing strategies are compared based
on the selection probability of $Z_1$, hence, the number of replications in which the true split 
variable $Z_1$ is detected with a $p$-value smaller than the predefined significance level 
$\alpha = 0.05$. This choice of measurement is due to the aim of investigating how well the 
tests perform as significance test.
However, ignoring this significance level yields the same conclusions which can be seen when 
comparing Figures~\ref{fig:scores_stump_wosign} and \ref{fig:scores_stump}, both being based 
on the exact same evaluation, but in Figure~\ref{fig:scores_stump_wosign} the proportion of 
replications in which the true splitting variable $Z_1$ is detected by showing the smallest $p$-value, 
but not necessarily smaller than the significance level, is illustrated.

\section{Continuous ``stump'' scenario}
\label{app:contbeta}
